\documentclass[11pt,a4paper]{amsart}

\usepackage[utf8]{inputenc}
\usepackage{amsmath,amsthm,amssymb}

\newtheorem{theorem}{Theorem}
\newtheorem{proposition}{Proposition}
\theoremstyle{remark}
\newtheorem{remark}{Remark}

\title{Monotone Derivatives are Continuous}
\author{}
\date{14 March 2026}

\begin{document}

\maketitle

\begin{abstract}
Derivatives need not be continuous, but they are far from arbitrary functions.
A classical theorem of Darboux states that every derivative has the Intermediate Value Property.
In this note we explain how this fact interacts with monotonicity:
if the derivative $f'$ is monotone on an interval, then $f'$ must actually be continuous there.
We give two proofs.
The first isolates the key idea very directly: a monotone discontinuity would force a jump, while Darboux's Theorem forbids jumps for derivatives.
The second proof uses one-sided limits together with the Mean Value Theorem.
\end{abstract}

\section{Introduction}

In elementary calculus, one often first meets derivatives through familiar smooth examples:
polynomials, trigonometric functions, exponentials, and so on.
In such cases the derivative is not only defined, but also continuous.
It is therefore natural at first to suspect that differentiability should automatically imply continuity of the derivative.

This is false.
A function may be differentiable everywhere and yet have a derivative that fails to be continuous at some point.
A standard example is
\[
f(x)=
\begin{cases}
x^2\sin(1/x), & x\neq 0,\\
0, & x=0.
\end{cases}
\]
This function is differentiable on $\mathbb{R}$, but its derivative is not continuous at $0$.

Nevertheless, derivatives still satisfy an important regularity property.
They cannot jump from one value to another while skipping intermediate values.
This is the content of Darboux's Theorem: every derivative has the Intermediate Value Property.

The purpose of this note is to show that if we add one more hypothesis, namely monotonicity, then this residual regularity becomes strong enough to force continuity.

\begin{theorem}\label{thm:main}
Let $f:I\to\mathbb{R}$ be differentiable on an open interval $I$.
If $f'$ is monotone on $I$, then $f'$ is continuous on $I$.
\end{theorem}

This is a very natural result.
A monotone function may certainly be discontinuous, but any discontinuity of a monotone function is of jump type.
On the other hand, a derivative cannot have a jump, because of Darboux's Theorem.
The theorem above follows from combining these two observations.

\section{Darboux's Theorem}

We begin with the classical theorem on which the first proof rests.

\begin{theorem}[Darboux]\label{thm:darboux}
Let $f$ be continuous on $[a,b]$ and differentiable on $(a,b)$.
Suppose $x_1,x_2\in(a,b)$ with $x_1<x_2$, and let $k$ be a real number between $f'(x_1)$ and $f'(x_2)$.
Then there exists $c\in(x_1,x_2)$ such that
\[
f'(c)=k.
\]
\end{theorem}

In other words, $f'$ has the Intermediate Value Property, even though $f'$ need not be continuous.

\begin{proof}
Assume for definiteness that
\[
f'(x_1)<k<f'(x_2).
\]
Define
\[
g(x)=f(x)-kx.
\]
Then $g$ is continuous on $[x_1,x_2]$ and differentiable on $(x_1,x_2)$.
Moreover,
\[
g'(x_1)=f'(x_1)-k<0,
\qquad
g'(x_2)=f'(x_2)-k>0.
\]

Since $g$ is continuous on the compact interval $[x_1,x_2]$, it attains a minimum at some point $c\in[x_1,x_2]$.
Because $g'(x_1)<0$, the minimum cannot occur at $x_1$.
Likewise, because $g'(x_2)>0$, the minimum cannot occur at $x_2$.
Hence the minimum occurs at some interior point $c\in(x_1,x_2)$.

By Fermat's theorem,
\[
g'(c)=0.
\]
Therefore
\[
f'(c)-k=0,
\]
so $f'(c)=k$.
\end{proof}

\section{A preliminary observation on monotone functions}

The following elementary fact about monotone functions is all we need.

\begin{proposition}\label{prop:monotone-limits}
Let $g$ be monotone increasing on an open interval $I$, and let $c\in I$.
Then the one-sided limits
\[
\lim_{x\to c^-}g(x)
\qquad\text{and}\qquad
\lim_{x\to c^+}g(x)
\]
exist.
Moreover, if we write
\[
A:=\lim_{x\to c^-}g(x),
\qquad
B:=\lim_{x\to c^+}g(x),
\]
then
\[
A\le g(c)\le B.
\]
\end{proposition}

\begin{proof}
Because $g$ is increasing, the set of values $g(x)$ for $x<c$ is bounded above by $g(c)$, so its supremum exists.
This supremum is exactly the left-hand limit:
\[
A=\sup\{g(x):x<c\}.
\]
Similarly, the set of values $g(x)$ for $x>c$ is bounded below by $g(c)$, so its infimum exists, and
\[
B=\inf\{g(x):x>c\}.
\]
Monotonicity also immediately gives
\[
g(x)\le g(c)\le g(y)
\qquad (x<c<y),
\]
and hence
\[
A\le g(c)\le B.
\]
\end{proof}

\begin{remark}
A monotone function need not be continuous.
However, its discontinuities are highly restricted: they can only be jumps.
Proposition \ref{prop:monotone-limits} is the precise local expression of that fact.
\end{remark}

\section{First proof of the main theorem}

We now prove Theorem \ref{thm:main} using Darboux's Theorem.

\begin{proof}[Proof of Theorem \ref{thm:main}: first method]
It suffices to treat the case in which $f'$ is monotone increasing.
The decreasing case follows by applying the same argument to $-f'$.

Fix $c\in I$.
By Proposition \ref{prop:monotone-limits}, the one-sided limits
\[
A:=\lim_{x\to c^-}f'(x),
\qquad
B:=\lim_{x\to c^+}f'(x)
\]
exist and satisfy
\[
A\le f'(c)\le B.
\]

Suppose, toward a contradiction, that $f'$ is not continuous at $c$.
Then at least one of the inequalities
\[
A<f'(c),
\qquad
f'(c)<B
\]
must be strict.
In particular,
\[
A<B.
\]

Choose a number $k$ such that
\[
A<k<B
\]
and also $k\ne f'(c)$.
Now choose points $x_1,x_2\in I$ with
\[
x_1<c<x_2.
\]
Since $f'$ is increasing, we have
\[
f'(x_1)\le A<k<B\le f'(x_2).
\]
By Darboux's Theorem, there exists some $d\in(x_1,x_2)$ such that
\[
f'(d)=k.
\]

But this is impossible.

If $d<c$, then monotonicity gives
\[
f'(d)\le A<k,
\]
contradicting $f'(d)=k$.

If $d>c$, then monotonicity gives
\[
f'(d)\ge B>k,
\]
again contradicting $f'(d)=k$.

If $d=c$, then
\[
f'(c)=k,
\]
which contradicts the way $k$ was chosen.

Thus our assumption was false.
Hence $f'$ is continuous at $c$.
Since $c\in I$ was arbitrary, $f'$ is continuous on all of $I$.
\end{proof}

\section{Second proof via the Mean Value Theorem}

The first proof captures the main conceptual point.
For completeness, we also give a second proof that uses the Mean Value Theorem more explicitly.

\begin{proof}[Proof of Theorem \ref{thm:main}: second method]
Again assume that $f'$ is monotone increasing, and fix $c\in I$.
Let
\[
A:=\lim_{x\to c^-}f'(x),
\qquad
B:=\lim_{x\to c^+}f'(x).
\]
We shall prove that
\[
A=f'(c)=B.
\]

First consider $h>0$ sufficiently small.
By the Mean Value Theorem, there exists $\xi_h\in(c,c+h)$ such that
\[
\frac{f(c+h)-f(c)}{h}=f'(\xi_h).
\]
Because $c<\xi_h<c+h$ and $f'$ is increasing, we have
\[
B\le f'(\xi_h)\le f'(c+h).
\]
Hence
\[
B\le \frac{f(c+h)-f(c)}{h}\le f'(c+h).
\]
Now let $h\to 0^+$.
The middle term tends to $f'(c)$ because $f$ is differentiable at $c$, while the right-hand term tends to $B$ by definition.
Therefore
\[
B\le f'(c)\le B,
\]
so
\[
f'(c)=B.
\]

Now consider $h<0$ with $|h|$ small.
Again by the Mean Value Theorem, there exists $\xi_h\in(c+h,c)$ such that
\[
\frac{f(c+h)-f(c)}{h}=f'(\xi_h).
\]
Since $c+h<\xi_h<c$ and $f'$ is increasing,
\[
f'(c+h)\le f'(\xi_h)\le A.
\]
Thus
\[
f'(c+h)\le \frac{f(c+h)-f(c)}{h}\le A.
\]
Letting $h\to 0^-$, we obtain
\[
A\le f'(c)\le A,
\]
and so
\[
f'(c)=A.
\]

Combining the two identities gives
\[
A=f'(c)=B.
\]
Hence $f'$ is continuous at $c$.
Since $c$ was arbitrary, $f'$ is continuous on $I$.
\end{proof}

\section{Concluding remark}

The theorem fits into a broader theme in real analysis:
derivatives may behave irregularly, but they still satisfy strong structural constraints.
Darboux's Theorem is one of the most important such constraints.
Once monotonicity is added, the remaining room for pathology disappears, and continuity follows.

One may also view the theorem as a useful bridge between two familiar facts:
monotone functions have one-sided limits and only jump-type discontinuities, while derivatives have the Intermediate Value Property and therefore cannot jump.
Together, these facts force a monotone derivative to be continuous.

\begin{thebibliography}{9}

\bibitem{abbott}
Stephen Abbott,
\emph{Understanding Analysis},
2nd ed.,
Springer, New York, 2015.

\bibitem{courant-john}
Richard Courant and Fritz John,
\emph{Introduction to Calculus and Analysis. Vol. I},
Springer, New York, 1999.

\bibitem{lang}
Serge Lang,
\emph{Undergraduate Analysis},
2nd ed.,
Springer, New York, 1997.

\bibitem{rudin}
Walter Rudin,
\emph{Principles of Mathematical Analysis},
3rd ed.,
McGraw--Hill, New York, 1976.

\end{thebibliography}

\end{document}
